---
title: "Masters Homework 3"
author: "Your Name Here"
date: "7/29/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
nyc <- read.csv("https://raw.githubusercontent.com/fredlapolla/RVilcekMasters2021/main/NYC_HANES_DIAB.csv")
```



How does R treat blank values?

What command can be used to identify where NAs are located in a column?

```{r}

```


How can you use indexing to create a subset without any NAs?

```{r}

```

How could you delete a column of data, for example, the 5th 
```{r}

```


What command would let you find a set of text in the cells of a column?

```{r}

```


Rename the HSQ1 variable to Health_Status

```{r}

```

Use the Mutate command to create a ratio of total cholesterol divided by HDL

```{r}

```



>- Why would one use a visualization for exploratory analysis?



>- What information about the nature of our data can a histogram tell us?



>- Make a histogram on your own on one column from the nyc data frame. Is it skewed? Which way? Try confirming with pastecs::stat.desc (hint set norm = TRUE as an argument).

```{r}

```

>- Make a QQPlot to assess the normality of one variable of your choosing. Is the distribution normal? How can you tell?

```{r}

```

>- Make a boxplot glucose levels by diabetes diagnosis:

```{r}


```

>- Make a scatterplot of total cholesterol by LDL. Does there appear to be a relationship?

```{r}

```

