---
title: "Homework1"
author: "Fred LaPolla"
date: "6/11/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Homework 1

```{r, echo = TRUE}
nyc <-read.csv("https://raw.githubusercontent.com/fredlapolla/RVilcekMasters2021/master/NYC_HANES_DIAB.csv")

```

>- Create a list of numbers 1 to 100 and name this list FirstHundred

```{r}

```

>- On your own, explore nyc and find the median and mean Cadmium levels`

```{r}

```


>- Find summary information about the total blood mercury levels in our NYC HANES data (hint: names(nyc) will call the names of each column). What is the max value, what is the minimum value.  

```{r}

```

>- Bonus/Challenge: What is the Standard Deviation and skewness of the blood mercury?

```{r}

```


>- Install a package and load, such as "ggplot2" or "dplyr"

```{r}

```


>- What is a directory? How can you look up the current directory in which you are working in R?

```{r}

```


>- How does one read in a CSV file of data into R?

```{r}

```

>- What two commands are needed to install and use a package in R?

```{r}

```

>- What are reasons one would choose to use R for data analysis?


