---
title: "Masters Homework 2"
author: "YOUR NAME HERE"
date: "7/22/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Homework

>- When you read in a .csv file into R, what class is the object that you read in (like nyc)?



>- What command can you use to determine what type of data is in a column?

```{r}

```


>- Provide a command to find the type of data in the DX_DBTS column of nyc. 

```{r}

```


>- Provide the same command the type of data but use the other way of indexing from what you did above. 

```{r}

```



>- What would happen if we had chosen the first 4 columns of NYC HANES and made a matrix? What would the class of the cells in this matrix be? Why?

```{r}

```


>- If we make a vector,vec<- c(3,9,4,7,8,9), and multiply it by WannaMatrix, what is the value in the third row, third column? 

```{r}
WannaMatrix <- nyc[5:10, 6:9]
WannaMatrix <- as.matrix(WannaMatrix,nrow = 6, ncol = 4)



```



>- What happens if instead we made a vector vec2 <- c(3,9,4,7,8) and multiply it by WannaMatrix, what is the third row, third column value? What is going on? 

```{r}

```


>- Look up the arguments for read.csv

```{r}

```


>- Use apply() to find the median value of CADMIUM, LEAD, MERCURY TOTAL, HDL, CHOLESTEROLTOTAL, GLUCOSESI, CREATININESI, and CREATININE. The values should be:

 CADMIUM              LEAD 
           0.2936            1.0576 
MERCURYTOTALBLOOD               HDL 
           1.6044           55.0000 
 CHOLESTEROLTOTAL         GLUCOSESI 
         183.0000            4.8800 
     CREATININESI        CREATININE 
          70.7200            0.8000

```{r}

```


>- Do the same using sapply()

```{r}

```


>- Use a for loop to find the median value of rows 16:19

```{r}

```


>- Make a for loop to find the standard deviation (hint sd() ) of CHOLESTEROLTOTAL, GLUCOSESI, CREATININESI, AND CREATININE

CHOLESTEROLTOTAL        GLUCOSESI     CREATININESI 
      38.6872845        1.6677120       26.5131524 
      CREATININE 
       0.2999403 

```{r}

```


>- Bonus/Challenge Create a vector in which to put the standard deviations 

```{r}

```


>- Use ifelse to create a new variable of high and normal A1C dichotomizing with 7 as a threshold for high levels

```{r}

```


>- Subset to only have females (gender == "Female")

```{r}

```


>- Create a subset of males and find the median total cholesterol level

```{r}

```


>- Try to create a summary of the mean A1C using two factors, GENDER and DX_DBTS

```{r}

```

Find the class of the first three columns of data. What are they?

```{r}

```

