---
title: 'BMIN-GA 1005 Programming for Data Analysis: Exploratory Data Visualization'
author: "Fred LaPolla"
date: "7/26/2020"
output: slidy_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Data Visualization

***

## Objectives

</br>
</br>
</br>

Students will be able to:

>- Explore their data to visually assess skew
>- Create a heatmap to visually assess gene expression levels
>- Use Ggplot2 to build different types of charts by employing layers 
>- Create charts and figures for publication including customzing colors and shapes, as well as exporting figures from R.

***

## Getting started

Let's pull in our same NYC HANES data set. 

```{r, echo = TRUE}
nyc <-read.csv("https://raw.githubusercontent.com/fredlapolla/RVilcekMasters2021/main/NYC_HANES_DIAB.csv")

nyc$AGEGROUP <- factor(nyc$AGEGROUP, levels = 1:3, labels = c("Youngest", "Middle", "Aged"))
nyc$GENDER <- factor(nyc$GENDER, levels = 1:2, labels = c("male", "female"))
# Rename the HSQ_1 factor for identification
  nyc$HSQ_1 <- factor(nyc$HSQ_1, levels = 1:5, labels=c("Excellent","Very Good","Good", "Fair", "Poor"))
  # Rename the DX_DBTS as a factor
  nyc$DX_DBTS <- factor(nyc$DX_DBTS,levels = 1:3, labels=c("Diabetes with DX","Diabetes with no DX","No Diabetes"))
  
```

There are two major reasons we might want to visualize our data: exploration and communication. 

***

## Data Exploration

</br>
</br>
</br>

In the first class, what were some of the ways we discussed to get a sense of our data? Why do we do this?

>- Summary
>- paste.ecs::stat.desc
>- psych::describe
>- Others



In addition to getting an understanding of our data, these tools help us understand what sort of tests are possible on our data type.



***

## Data Exploration

</br>
</br>
</br>

One of the first things we may want to know is "Is my data normally distributed"? Why might we want to know this?

>- Different statistical tests for hypothesis testing assume a Gaussian or Normal Distribution

>- A quick way to get a sense of the nature of our distribution is to look at a picture, and a common option is a **Histogram**

***

## Data Exploration


The command hist() let's us make a histogram:

```{r, echo = TRUE}
#For today don't worry about par, this is to plot three charts side by side
threechart <-par(mfrow = c(1,3))
#this is the meat of the matter
hist(nyc$GLUCOSE)

hist(nyc$CHOLESTEROLTOTAL)

hist(nyc$LEAD)
par(mfrow = c(1,1))
```

***

## Hist colors


You could also give it a fill if you want:

```{r, echo = TRUE}
hist(nyc$CHOLESTEROLTOTAL, col = "steelblue")
```

***

## Group Work

Try to make a histogram of MERCURYU and LDL Estimate. Are these  normally distributed or skewed, and if so, which way?

```{r}

```




***

## Boxplot



We can also use boxplots, sometimes also called Box and Whisker or Tukey plots. 

```{r, echo = TRUE}
boxplot(nyc$CHOLESTEROLTOTAL)
```

Boxplots display: the median, quartiles and outliers (which are by default 1.5 x the interquartile range above or below the quartiles). You can tell if there is relatively a lot of skew or not because the median will be near the mean. 

*** 

## Boxplots



You can also use boxplots to look at comparisons of groups. 

```{r, echo = TRUE}
boxplot(nyc$CHOLESTEROLTOTAL ~ nyc$GENDER)
```

***

## Boxplots

You can also add in a notch, which if the notch overlaps is a good rule of thumb that they are not statistically significantly different (but beware of statistical significance).

```{r, echo = TRUE}
boxplot(nyc$CHOLESTEROLTOTAL ~ nyc$GENDER, notch = TRUE)
```

***

## Boxplots

With a color

```{r, echo = TRUE}
boxplot(nyc$CHOLESTEROLTOTAL ~ nyc$GENDER, notch = TRUE, col = "pink")
```

*** 


## Group Work

Make a box plot of A1C by diabetes diagnosis. Are they likely the same for both groups?

```{r}

```

*** 

## QQ Plots


Another option for looking at skew is the Quantile-Quantile Plot, or QQ Plot. This chart plots quantiles, similar to percentiles of our data against a normally distributed quantile. Baiscally, the more our data lies on a straight line, the more normal it is. Compare:

```{r, echo=TRUE}
#QQNorm plots how the quantiles of our sample compare to a theoretical normal distribution
par(mfrow=c(1,2))
##QQPlot for total cholesterol
qqnorm(nyc$CHOLESTEROLTOTAL, main = "Normal Q-Q Plot Total Cholesterol")
#QQLine adds a straight line for reference
qqline(nyc$CHOLESTEROLTOTAL)

##QQPlot for Cadmium
qqnorm(nyc$CADMIUM,main= "Normal Q-Q Plot Cadmium")
qqline(nyc$CADMIUM)
```



## Group Work

Make a qq plot of LDL. Also make one for MercuryU. Would you expect either of these to be normal or skewed? How can you tell?

```{r}

```




***

## Correlations & Scatter Plots

</br>
</br>
</br>

Sometimes we may be running a correlation test. While running Pearson's or a Mann-Whitney Correlation test can give you results, it can also be nice to look at an image of that correlation to see how nicely (or not) the points seem to correlate:

```{r, echo = TRUE}
plot(nyc$A1C, nyc$GLUCOSE)
cor.test(nyc$A1C, nyc$GLUCOSE, method = "spearman" , alternative = "two.sided")
```


## Correlation & Scatter Plots

</br>
</br>
</br>

Sometimes, for example in cases where our scale has exponentially larger outliers, it can be helpful to try to normalize our data so that the chart is more viewable. You could try a transformation like "log()". It is helpful to do the log +1 in this case because log(0) is negative infinity

```{r, echo = TRUE}
par(mfrow = c(1,2))
plot(nyc$GLUCOSE, nyc$CHOLESTEROLTOTAL)
plot(log(nyc$GLUCOSE+1), log(nyc$CHOLESTEROLTOTAL+1))
```

***

## Group Work

If we wanted to make a scatter plot of lead by cotinine, how would we do it? Based on what you now know about assessing skew, what test might we run for this?

```{r}

```


